---
title: Basic Mapping
author: Daniel Redwine
output: 
  html_document:
    toc: True # table of content true
    toc_depth: 2  # up to one depth of headings, no subheader (specified by # in html)
    toc_float: true
    collapsed: no
    smooth_scroll: yes
    highlight: pygments
    df_print: paged
    number_sections: false  ## if you want number sections at each table header
    theme: cosmo  # many options for theme
---
```{r setup, include=FALSE}

#install.packages('pacman')
#pacman::p_load("leaflet","OpenStreetMap","sf","tidyverse") #to install needed packages

# Load libraries
library(pacman)
library(leaflet)
library(OpenStreetMap)
library(sf)
library(tidyverse)
library(dplyr)
```
# Campus Spot Mapping

Ornithology students in Spring 2023 completed a project where Eastern Bluebirds(EABL), Northern Mockingbirds(NOMO), Eurasian Starlings(EUST), and House Sparrows(HOSP) were spot mapped across the APSU campus. EABL and NOMO are native species, while EUST and HOSP are invasive species. This allowed analysis of densities for both individual species as well as native versus invasive species across campus. However, recent updates to R packages has rendered that script useless. Basic maps for the spot mapping project will be produced here using currently available packages.

```{r Load and clean dataset, echo=FALSE, message=FALSE, warning=FALSE}

spotmap_data <- read.csv("../data/2023data.csv") # load in dataset, .. moves up one level

spotmap_data$lat<-as.numeric(spotmap_data$lat) # make sure lat and lon are numeric 
spotmap_data$lon<-as.numeric(spotmap_data$lon)
spotmap_data$species<-as.factor(spotmap_data$species) # species as factor
```
## Static Map
```{r static map, echo=FALSE, message=FALSE, warning=FALSE, fig.show = "hold", out.width = "100%", fig.align = "center", fig.cap="Figure 1. Map showing all locations for EABL, NOMO, EUST, and HOSP mapped for the 2023 ornithology spot mapping project."}

# Set the extent (max/min lat/lon) and open the base map
static_base_map <- openmap(c(max(spotmap_data$lat)+0.001, min(spotmap_data$lon)-0.001), + 
              c(min(spotmap_data$lat)-0.001, max(spotmap_data$lon)+0.001), type = "osm")

# Projection onto the map 
spotmap_projection <- openproj(static_base_map, projection = "+proj=longlat +datum=WGS84")

autoplot.OpenStreetMap(spotmap_projection, expand = TRUE) + 
  geom_point(data=spotmap_data, aes(lon, lat)) +
  geom_point(data = spotmap_data, aes(lon, lat), size = 2) + 
  aes(color = species) +
  theme(axis.title = element_text(face="bold")) + 
  labs(x="Longitude", y="Latitude") +
  guides(color=guide_legend("Species")) +  theme_bw()
```
## Interactive Map 
```{r interactive map, echo=FALSE, message=FALSE, warning=FALSE, fig.show = "hold", out.width = "100%", fig.align = "center", fig.cap="Figure 2. Interactive map showing all locations for EABL, NOMO, EUST, and HOSP mapped for the 2023 ornithology spot mapping project."}

HOSP_data <- spotmap_data %>%
  filter(species == "HOSP")

NOMO_data <- spotmap_data %>%
  filter(species == "NOMO")

EUST_data <- spotmap_data %>%
  filter(species == "EUST")

EABL_data <- spotmap_data %>%
  filter(species == "EABL")

leaflet() %>% 
  addTiles(group = "OSM")%>%
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB") %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap, group = "NatGeo") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "ESRI") %>%
  setView(lng = -87.356090, lat = 36.536090, zoom = 15) %>%
  addCircleMarkers(popup = HOSP_data$num_ind,
                   label = HOSP_data$species,
                   group = "House Sparrow",
                   lng = HOSP_data$lon, 
                   lat = HOSP_data$lat,
                   weight = 1,
                   radius = 6, 
                   color = "black",
                   fillColor = "brown",
                   fillOpacity = 0.7) %>%
  addCircleMarkers(popup = EABL_data$num_ind,
                   label = EABL_data$species,
                   group = "Eastern Bluebird",
                   lng = EABL_data$lon, 
                   lat = EABL_data$lat,
                   radius = 6, 
                   weight = 1,
                   color = "black",
                   fillColor = "blue",
                   fillOpacity = 0.7) %>%
    addCircleMarkers(popup = NOMO_data$num_ind,
                   label = NOMO_data$species,
                   group = "Northern Mockingbird",
                   lng = NOMO_data$lon, 
                   lat = NOMO_data$lat,
                   radius = 6, 
                   weight = 1,
                   color = "black",
                   fillColor = "grey",
                   fillOpacity = 0.7) %>%
    addCircleMarkers(popup = EUST_data$num_ind,
                   label = EUST_data$species,
                   group = "Eurasian Starling",
                   lng = EUST_data$lon, 
                   lat = EUST_data$lat,
                   radius = 6, 
                   weight = 1,
                   color = "black",
                   fillColor = "purple",
                   fillOpacity = 0.7) %>%
    addLayersControl(
    baseGroups = c("OSM", "CartoDB", "NatGeo", "ESRI"),
    options = layersControlOptions(collapsed = FALSE),
    overlayGroups = c("House Sparrow", "Eastern Bluebird", "Northern Mockingbird", "Eurasian Starling"))
```